Estimating process-based model parameters from species distribution data using the evolutionary algorithm CMA-ES

### 2023-04-18 manuscript/paper.tex
removed green annotations for reviewers

\noindent \begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Two main types of species distribution models are used to project
  species range shifts in future climatic conditions: correlative and
  process-based models. Although there is some continuity between these
  two types of models, they are fundamentally different in their
  hypotheses (statistical relationships \emph{vs} mechanistic
  relationships) and their calibration methods
  \textcolor{darkgreen}{(SDMs tend to be occurrence data-driven while PBMs tend to be prior-driven)}.
\item
  One of the limitations to the use of process-based models is the
  difficulty to parameterize them for a large number of species
  \textcolor{darkgreen}{compared to correlative SDMs}. We investigated
  the feasibility of using an evolutionary algorithm (called covariance
  matrix adaptation evolution strategy, CMA-ES) to calibrate
  process-based models using species distribution data. This method is
  well established in some fields (robotics, aerospace research,
  \ldots), but has never been used, to our knowledge, in ecology,
  despite its ability to deal with very large space dimensions. Using
  tree species occurrence data across Europe, we adapted the CMA-ES
  algorithm to find appropriate values of model parameters. We estimated
  simultaneously 27 to 77 parameters of two process-based models
  simulating forest tree's ecophysiology for three species with varying
  range sizes and geographical distributions.
\item
  CMA-ES provided parameter estimates leading to better prediction of
  species distribution than parameter estimates based on experts
  knowledge. Our results also revealed that some model parameters and
  processes were strongly dependent, and different parameter
  combinations could therefore lead to high model accuracy.
\item
  \textcolor{darkgreen}{We conclude that} CMA-ES is an efficient
  state-of-the-art method to calibrate process-based models with a large
  number of parameters using species occurrence data. Inverse modelling
  using CMA-ES is a powerful method to calibrate process-based
  parameters which can hardly be measured. However, the method does not
  warranty that parameter estimates are correct because of several
  sources of bias, similarly to correlative models, and expert knowledge
  is required to validate results.
\end{enumerate}

@@

Calibration (i.e.~parameter setting and estimation) is a fundamental
step in the modelling process. It allows the model to reproduce the
reality with more or less success. The result of the calibration
provides insights on the ability of the model to reproduce and explain
reality (model predictive power). Calibration of complex models such as
process-based SDMs is time-consuming, and modelers are often challenged
by the dimension of the parameter space, the complexity of the possible
correlations among parameters, and the scarcity of observed experimental
data to calibrate them. Therefore, although the philosophy of such
models is to measure parameters, statistical inference might be useful
when data are not yet available to infer parameter values. Parameter
inference can be achieved through several methods which have been
developed in the last decades.
\textbackslash textcolor\{darkgreen\}\{Most of them fall into two
categories: informal and statistical calibration. On the one hand,
statistical calibration assumes a data-generating model, and a
likelihood function, which quantifies the probability of the observed
data given the model parameters, is expressed mathematically. This
likelihood is the foundation of \emph{maximum likelihood estimation} and
\emph{Bayesian inference}, two commonly used methods for parameter
estimation. In practice, deriving the likelihood function can be quite
challenging for complex models with many parameters and thresholding.
One approach is to use numerical methods such as
\emph{simulation-based inference} or
\emph{approximate Bayesian computation} (e.g.
\protect\hyperlink{ref-Hartig2014}{Hartig \emph{et al.}
2014})\textcolor{darkgreen}{. However, obtaining a reliable estimate of the likelihood of a complex process-based model can require a large number of simulated samples, which can be computationally expensive. On the other hand, informal calibration uses an informal objective function to measure the discrepancy between the model predictions and the observed data. By minimizing the objective function, one is able to identify the parameter values that best fit the observed data, even though it may not have the same statistical rigor as statistical calibration. This optimization step can be achieved with several methods, either deterministic (e.g. Nelder-Mead method) or stochastic (e.g. simulated annealing, evolutionary algorithms).}\\
Recently, an approach belonging to the evolutionary algorithm family,
called Covariance Matrix Adaptation Evolution Strategy (CMA-ES), has
been proposed (\protect\hyperlink{ref-Hansen2001}{Hansen \& Ostermeier
2001}). One of the advantages of CMA-ES is its ability to cumulate
information over iterations in order to adapt its own parameters (in
particular the covariance matrix), which makes it more robust to noise.
CMA-ES is especially performant for non-separable problems (i.e.~when
the model parameters are dependent) and large search space. This method
has been successfully applied in various fields such as aerospace (e.g.
\protect\hyperlink{ref-Collange2010}{Collange \emph{et al.} 2010}),
optics (e.g. \protect\hyperlink{ref-Gagne2008}{Gagné \emph{et al.}
2008}), and robotics (e.g. \protect\hyperlink{ref-Hill2020}{Hill
\emph{et al.} 2020}). CMA-ES is acknowledged to be one of the most
efficient approaches in continuous black-box optimization
(\protect\hyperlink{ref-Hansen2010}{Hansen \emph{et al.} 2010}) but to
our knowledge has never been used in ecology.

@@

Here we demonstrated that inverse calibration with CMA-ES is feasible
and provide good results for complex and runtime-expensive ecological
models.\\
With a subsampling of species occurrence data, the algorithm succeeds in
finding parameter sets which provide \textcolor{darkgreen}{higher} AUC
values. The predictions of the calibrated models are sharply improved
compared to expert parametrization
(\hyperref[fig:phenofitmaps]{Figure 1} and
\hyperref[fig:castaneamaps]{Figure 2}). Two striking examples are the
increase in the performance of PHENOFIT model for silver fir, from 0.72
to 0.9, and of CASTANEA model for holm oak, from 0.7 to 0.95. Moreover,
CMA-ES performed equally well regardless of the species occurrence
subset used during calibration
(\hyperref[fig:cmaesrepAUCcal]{Figure 3.B.}), and thus permitted to find
a good compromise between computational cost and calibration efficiency.

CMA-ES is a ``generic'' optimizer which can be applied to various
problems. It is easy to use as it does not require an extensive tuning
to efficiently explore the parameter space. We only had to choose the
number of candidate solutions \(\lambda\), and the initial search region
(initial starting point and step size \(\sigma\)). As well as being
quasi parameter-free, CMA-ES has several structural advantages
particularly useful for complex optimization problems.
\textcolor{darkgreen}{First, the algorithm's covariance matrix enables the learning of second-order information, which provides insights into pairwise dependencies between parameters. Second, the covariance matrix adaptation of CMA-ES is highly efficient in handling ill-conditioned and non-separable problems }(\protect\hyperlink{ref-Hansen2011}{Hansen
\emph{et al.}
2011})\textcolor{darkgreen}{. Last, CMA-ES's update mechanism of step size $\sigma$ (i.e. the mutation force) helps prevent premature convergence }(\protect\hyperlink{ref-Hansen2001}{Hansen
\& Ostermeier
2001})\textcolor{darkgreen}{, allowing the algorithm to explore more of the search space.}
CMA-ES has been shown to outperform several other optimization
algorithms (\protect\hyperlink{ref-Hansen2010}{Hansen \emph{et al.}
2010}), and is usually the most efficient method when the target cost
(i.e.~the number of objective function evaluations) is about \(100*N\)
(\(N\) being the dimension of the parameter search space,
\protect\hyperlink{ref-Baeck2013}{Bäck \emph{et al.} 2013}).

@@

Consequently, several parameter sets may be statistically equivalent and
parameters non-identifiable. In fact, calibration repetitions gave
diverging parameter values (\hyperref[fig:unfoldingplots]{Figure 4})
while being efficient in distinguishing between species presence and
absence (i.e.~AUC \textasciitilde{} 0.9,
\hyperref[fig:cmaesrepAUCcal]{Figure 3}). Thus, even if the calibrated
model describes the observed species distribution very well, it does not
necessarily mean that parameter values are ecologically relevant. This
concern is similar to the criticisms against correlative SDMs, in which
parameter values and correlations that well reproduce species ranges do
not necessarily describe a complex biological reality. In our case, the
constraints imposed by the explicit mathematical equations embedded in
the models we used were not sufficient to ensure calibration convergence
towards similar solutions that would have suggested a high biological
realism. However, it is worth noting that we deliberately chose large
parameter ranges (although biologically realistic, i.e.~corresponding to
the observations made on the different processes modelled across
different species) in order to give free rein to the optimization
algorithm. As our goal was to assess the performance of CMA-ES
objectively, we did not attempt to minimize this non-identifiability
issue by restricting the parameter space. To deal with equifinality
issues, an avenue to explore could be the use of multiple objective
functions during model calibration to assess different aspects of the
model performance.
\textcolor{darkgreen}{Additionally, if a closed-form likelihood can be derived, one could use a Bayesian framework to combine prior knowledge and inverse estimation of parameters to constrain the parameter space and study the nature of trade-offs between parameters}
(\protect\hyperlink{ref-Hartig2012}{Hartig \emph{et al.} 2012};
\protect\hyperlink{ref-Cailleret2020}{Cailleret \emph{et al.} 2020}).

@@

Our goal here was to investigate the performance of CMA-ES to calibrate
quite complex process-based species distribution models using species
occurrence data. We did not attempt to validate our parametrizations
using temporally or spatially independent data, and AUC was only used to
determine if model outputs were consistent with species distributions.
\textcolor{darkgreen}{However, AUC is scale-invariant: it measures how well predictions are ranked rather than their absolute values. For example, with PHENOFIT, a species with a fitness of $0.8$ could be considered as absent while another one with the same fitness could be considered as present. Therefore, when it is used as an objective function for model calibration, we probably lack some precision and consistency among species’ parameters estimation.}
Further work could thus be conducted to examine the effects of choosing
a different objective function.

>>>>

\noindent \begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Two main types of species distribution models are used to project
  species range shifts in future climatic conditions: correlative and
  process-based models. Although there is some continuity between these
  two types of models, they are fundamentally different in their
  hypotheses (statistical relationships \emph{vs} mechanistic
  relationships) and their calibration methods (SDMs tend to be
  occurrence data-driven while PBMs tend to be prior-driven).
\item
  One of the limitations to the use of process-based models is the
  difficulty to parameterize them for a large number of species compared
  to correlative SDMs. We investigated the feasibility of using an
  evolutionary algorithm (called covariance matrix adaptation evolution
  strategy, CMA-ES) to calibrate process-based models using species
  distribution data. This method is well established in some fields
  (robotics, aerospace research, \ldots), but has never been used, to
  our knowledge, in ecology, despite its ability to deal with very large
  space dimensions. Using tree species occurrence data across Europe, we
  adapted the CMA-ES algorithm to find appropriate values of model
  parameters. We estimated simultaneously 27 to 77 parameters of two
  process-based models simulating forest tree's ecophysiology for three
  species with varying range sizes and geographical distributions.
\item
  CMA-ES provided parameter estimates leading to better prediction of
  species distribution than parameter estimates based on experts
  knowledge. Our results also revealed that some model parameters and
  processes were strongly dependent, and different parameter
  combinations could therefore lead to high model accuracy.
\item
  We conclude that CMA-ES is an efficient state-of-the-art method to
  calibrate process-based models with a large number of parameters using
  species occurrence data. Inverse modelling using CMA-ES is a powerful
  method to calibrate process-based parameters which can hardly be
  measured. However, the method does not warranty that parameter
  estimates are correct because of several sources of bias, similarly to
  correlative models, and expert knowledge is required to validate
  results.
\end{enumerate}

@@

Calibration (i.e.~parameter setting and estimation) is a fundamental
step in the modelling process. It allows the model to reproduce the
reality with more or less success. The result of the calibration
provides insights on the ability of the model to reproduce and explain
reality (model predictive power). Calibration of complex models such as
process-based SDMs is time-consuming, and modelers are often challenged
by the dimension of the parameter space, the complexity of the possible
correlations among parameters, and the scarcity of observed experimental
data to calibrate them. Therefore, although the philosophy of such
models is to measure parameters, statistical inference might be useful
when data are not yet available to infer parameter values. Parameter
inference can be achieved through several methods which have been
developed in the last decades. Most of them fall into two categories:
informal and statistical calibration. On the one hand, statistical
calibration assumes a data-generating model, and a likelihood function,
which quantifies the probability of the observed data given the model
parameters, is expressed mathematically. This likelihood is the
foundation of \emph{maximum likelihood estimation} and
\emph{Bayesian inference}, two commonly used methods for parameter
estimation. In practice, deriving the likelihood function can be quite
challenging for complex models with many parameters and thresholding.
One approach is to use numerical methods such as
\emph{simulation-based inference} or \textbackslash emph\{approximate
Bayesian computation (e.g. \protect\hyperlink{ref-Hartig2014}{Hartig
\emph{et al.} 2014}). However, obtaining a reliable estimate of the
likelihood of a complex process-based model can require a large number
of simulated samples, which can be computationally expensive. On the
other hand, informal calibration uses an informal objective function to
measure the discrepancy between the model predictions and the observed
data. By minimizing the objective function, one is able to identify the
parameter values that best fit the observed data, even though it may not
have the same statistical rigor as statistical calibration. This
optimization step can be achieved with several methods, either
deterministic (e.g.~Nelder-Mead method) or stochastic (e.g.~simulated
annealing, evolutionary algorithms).\\
Recently, an approach belonging to the evolutionary algorithm family,
called Covariance Matrix Adaptation Evolution Strategy (CMA-ES), has
been proposed (\protect\hyperlink{ref-Hansen2001}{Hansen \& Ostermeier
2001}). One of the advantages of CMA-ES is its ability to cumulate
information over iterations in order to adapt its own parameters (in
particular the covariance matrix), which makes it more robust to noise.
CMA-ES is especially performant for non-separable problems (i.e.~when
the model parameters are dependent) and large search space. This method
has been successfully applied in various fields such as aerospace (e.g.
\protect\hyperlink{ref-Collange2010}{Collange \emph{et al.} 2010}),
optics (e.g. \protect\hyperlink{ref-Gagne2008}{Gagné \emph{et al.}
2008}), and robotics (e.g. \protect\hyperlink{ref-Hill2020}{Hill
\emph{et al.} 2020}). CMA-ES is acknowledged to be one of the most
efficient approaches in continuous black-box optimization
(\protect\hyperlink{ref-Hansen2010}{Hansen \emph{et al.} 2010}) but to
our knowledge has never been used in ecology.

@@

Here we demonstrated that inverse calibration with CMA-ES is feasible
and provide good results for complex and runtime-expensive ecological
models.\\
With a subsampling of species occurrence data, the algorithm succeeds in
finding parameter sets which provide higher AUC values. The predictions
of the calibrated models are sharply improved compared to expert
parametrization (\hyperref[fig:phenofitmaps]{Figure 1} and
\hyperref[fig:castaneamaps]{Figure 2}). Two striking examples are the
increase in the performance of PHENOFIT model for silver fir, from 0.72
to 0.9, and of CASTANEA model for holm oak, from 0.7 to 0.95. Moreover,
CMA-ES performed equally well regardless of the species occurrence
subset used during calibration
(\hyperref[fig:cmaesrepAUCcal]{Figure 3.B.}), and thus permitted to find
a good compromise between computational cost and calibration efficiency.

CMA-ES is a ``generic'' optimizer which can be applied to various
problems. It is easy to use as it does not require an extensive tuning
to efficiently explore the parameter space. We only had to choose the
number of candidate solutions \(\lambda\), and the initial search region
(initial starting point and step size \(\sigma\)). As well as being
quasi parameter-free, CMA-ES has several structural advantages
particularly useful for complex optimization problems. First, the
algorithm's covariance matrix enables the learning of second-order
information, which provides insights into pairwise dependencies between
parameters. Second, the covariance matrix adaptation of CMA-ES is highly
efficient in handling ill-conditioned and non-separable problems
(\protect\hyperlink{ref-Hansen2011}{Hansen \emph{et al.} 2011}). Last,
CMA-ES's update mechanism of step size \(\sigma\) (i.e.~the mutation
force) helps prevent premature convergence
(\protect\hyperlink{ref-Hansen2001}{Hansen \& Ostermeier 2001}),
allowing the algorithm to explore more of the search space. CMA-ES has
been shown to outperform several other optimization algorithms
(\protect\hyperlink{ref-Hansen2010}{Hansen \emph{et al.} 2010}), and is
usually the most efficient method when the target cost (i.e.~the number
of objective function evaluations) is about \(100*N\) (\(N\) being the
dimension of the parameter search space,
\protect\hyperlink{ref-Baeck2013}{Bäck \emph{et al.} 2013}).

@@

Consequently, several parameter sets may be statistically equivalent and
parameters non-identifiable. In fact, calibration repetitions gave
diverging parameter values (\hyperref[fig:unfoldingplots]{Figure 4})
while being efficient in distinguishing between species presence and
absence (i.e.~AUC \textasciitilde{} 0.9,
\hyperref[fig:cmaesrepAUCcal]{Figure 3}). Thus, even if the calibrated
model describes the observed species distribution very well, it does not
necessarily mean that parameter values are ecologically relevant. This
concern is similar to the criticisms against correlative SDMs, in which
parameter values and correlations that well reproduce species ranges do
not necessarily describe a complex biological reality. In our case, the
constraints imposed by the explicit mathematical equations embedded in
the models we used were not sufficient to ensure calibration convergence
towards similar solutions that would have suggested a high biological
realism. However, it is worth noting that we deliberately chose large
parameter ranges (although biologically realistic, i.e.~corresponding to
the observations made on the different processes modelled across
different species) in order to give free rein to the optimization
algorithm. As our goal was to assess the performance of CMA-ES
objectively, we did not attempt to minimize this non-identifiability
issue by restricting the parameter space. To deal with equifinality
issues, an avenue to explore could be the use of multiple objective
functions during model calibration to assess different aspects of the
model performance. Additionally, if a closed-form likelihood can be
derived, one could use a Bayesian framework to combine prior knowledge
and inverse estimation of parameters to constrain the parameter space
and study the nature of trade-offs between parameters
(\protect\hyperlink{ref-Hartig2012}{Hartig \emph{et al.} 2012};
\protect\hyperlink{ref-Cailleret2020}{Cailleret \emph{et al.} 2020}).

@@

Our goal here was to investigate the performance of CMA-ES to calibrate
quite complex process-based species distribution models using species
occurrence data. We did not attempt to validate our parametrizations
using temporally or spatially independent data, and AUC was only used to
determine if model outputs were consistent with species distributions.
However, AUC is scale-invariant: it measures how well predictions are
ranked rather than their absolute values. For example, with PHENOFIT, a
species with a fitness of \(0.8\) could be considered as absent while
another one with the same fitness could be considered as present.
Therefore, when it is used as an objective function for model
calibration, we probably lack some precision and consistency among
species' parameters estimation. Further work could thus be conducted to
examine the effects of choosing a different objective function.
